Making Q-learning robust to time discretization

### 2019-01-17 paper/icml_drau/framework.tex
.

\subsection{Return, discount factor  and Value function in near continuous time}
\label{sec:ret-gamma-v}

	In what follows, $\deltat$-dependant definitions of the discount factor $\gamma_\deltat$
	and the reward $r_\deltat$ are given. These definitions fit the discrete case
	scenario when $\deltat=1$, and yields well defined, non trivial returns
	and value functions when $\deltat$ goes to $0$.

\paragraph{Return and Discount Factor}
For a continuous MDP and a continuous trajectory $\tau = (s_t, a_t)_t$,
the return is defined as~\cite{cont_rl}
\begin{equation}
	\label{eq:continuous-return}
	R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt
\end{equation}

The natural discretization of this equation, if we want to define a discretized
return $R_\delta$ for each $\deltat$ is:
\begin{equation}
	\label{eq:discretized-return}
	R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
	r(s^\deltat_k, a^\deltat_k)\deltat
\end{equation}

@@

	The notion of time horizon, which is, informally, the amount of time on
	which the agent optimizes its return, is only meaningful in physical
	time.

        %Having a time horizon of $10$ steps means something very different
	%if $\deltat = 0.1$ and if $\deltat = 0.0001$.

        A rule of thumb
	is that the time horizon of an agent with discount factor $\gamma$ is
	of order $\frac{1}{1 - \gamma}$ steps. If $\gamma_\deltat$ was left constant
	as $\deltat$ goes to $0$, the corresponding time horizon in physical time
	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
	$\deltat$ goes to $0$.
        On the contrary, the definition of $\gamma_\deltat$ in
	Eq.~\eqref{eq:def-gamma} yields an approximately constant
	time horizon in physical time when $\deltat$ goes to $0$. Indeed,
	\begin{equation}
          \label{eq:time-horizon}
          \frac{\deltat}{1-\gamma^\deltat}
          = - \frac{1}{\log \gamma} + \bigO(\deltat)
          \approx \frac{1}{1-\gamma},
        \end{equation}
        approximately invariant to $\deltat$ when $\deltat$ goes to 0.

@@

The definition of the return Eq.~\eqref{eq:continuous-return}, yields the
following definition of the continuous value function
\begin{align}
  V^\pi(s) =& \E_{\tau\sim\pi}\left[R(\tau) \mid s_0 = s\right]\\
  =& \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t}\deltat r_{t} \mid s_0 = s\right].
\end{align}
The corresponding discretization for time discretization $\deltat$ is
\begin{align}
  V^\pi_\deltat(s) =& \E_{\tau\sim\pi}\left[R_\deltat(\tau) \mid s_0 = s\right]\\
  =& \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}\deltat r_{k\deltat} \mid s_0 = s\right]
\end{align}
which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
$r_\deltat$ defined in~(\ref{eq:def-gamma}, \ref{eq:def-reward}).
We have the following convergence theorem of the discretized value function to the continuous value function:
  \begin{theorem}
   For all $s \in {\cal S}$:
    \begin{equation}
      \label{eq:conv-value}
      V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)
    \end{equation}
  \end{theorem}

The discretized value function follows the Bellman equation:
\begin{equation}
  \label{eq:bellman}
  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\deltat} \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
\end{equation}

If the MDP is following the Eq. \eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$:\TODO{Cite ?}
\begin{equation}
  \label{eq:hamilton-jacobi-bellman}
  r + \partial_s V^\pi \cdot F + \frac{1}{2} \text{tr}\left(\Sigma^T\partial^2_{s^2} V^\pi\Sigma\right) = - \log(\gamma) V^\pi(s)
\end{equation}

@@

Contrary to the Value function, the Action-Value function $Q$ is ill-defined
for a continuous MDP. More precisely, the Q-function  collapses to the V-function when $\deltat \rightarrow 0$. In near continuous time,
the effect of individual actions on the Q-function is of order $\bigO(\deltat)$. This makes deriving a hierarchy of action from an approximate Q-function difficult.
Formally:
\begin{theorem}
  For all $s, a$, we have
  \begin{equation}
    \label{eq:q-discr}
    Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
  \end{equation}

  If we consider the continuous action-value function $Q^\pi$, we have:
  \begin{equation}
    \label{eq:q-cont}
    Q^\pi(s, a) = V^\pi
  \end{equation}
  \label{th:q-cont}
\end{theorem}
% Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
% $\deltat \ll~1$, Ito's lemma yield
% \begin{align}
% 	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
% 		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
% 		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
% 		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
% 		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
% 		    \label{eq:Q_cont}
                      %   \end{align}

A consequence of Theorem \ref{th:q-cont} is that in exact continuous time,
$Q^\pi$ does not bear any information on the hierarchy of actions, and
thus cannot be used to select actions that yields higher future returns. There is no time continuous Q-learning.

\begin{proof}
  \begin{align}
    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V^\pi_\deltat(s')\right] 
  \end{align}
  But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
  \begin{equation}
    \E_{s'|s, a}\left[\delta s\right] = \bigO(\deltat) \quad \text{Var}(\delta s) = \bigO(\deltat)
  \end{equation}
  \TODO{Explain. And this is not exactly the Variance.} Therefore:
  \begin{align}
    |\E_{s'|s, a}\left[V^\pi_\deltat(s')\right] - V^\pi_\deltat(s)| &= \E_{\delta s|s, a}\left[\partial_s V^\pi_\deltat\cdot \delta s +\bigO(\|\delta s\|^2)\right]\nonumber\\
                                            &\leq \|\partial_s V_\deltat^\pi\|\cdot \|\E_{\delta s|s, a}\left[\delta s\right]\| +\bigO(\delta t)\nonumber\\
    &= \bigO(\deltat)\nonumber
  \end{align}
  Finally:
  \begin{align}
    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V_\deltat^\pi(s')\right] \nonumber\\
                &= r(s, a)\deltat + \left(1 + \bigO(\deltat)\right)(V^\pi(s) + \bigO(\deltat))\nonumber\\
    &= V^\pi(s) + \bigO(\deltat)\nonumber
  \end{align}
  \end{proof}

>>>>

\subsection{Return, discount factor and value function in near continuous time}
\label{sec:ret-gamma-v}

	In what follows, $\deltat$-dependant scalings of the discount factor $\gamma_\deltat$
	and the reward $r_\deltat$ are given. These definitions fit the discrete case
	scenario when $\deltat=1$, and yields well defined, non trivial returns
	and value functions when $\deltat$ goes to $0$.

\paragraph{Return and Discount Factor}
For a continuous MDP and a continuous trajectory $\tau = (s_t, a_t)_t$,
the return is defined as~\cite{cont_rl}
\begin{equation}
	\label{eq:continuous-return}
	R(\tau) = \int_{t=0}^\infty\gamma^tr(s_t, a_t)dt.
\end{equation}

The natural discretization of this equation, if we want to define a discretized
return $R_\delta$ for each $\deltat$ is:
\begin{equation}
	\label{eq:discretized-return}
	R_\deltat(\tau) = \sum_{k=0}^\infty\gamma^{k\deltat}
	r(s_\deltat^k, a_\deltat^k)\deltat
\end{equation}

@@

	The notion of time horizon, which is, informally, the amount of time on
	which the agent optimizes its return, is only meaningful in physical
	time.
        %Having a time horizon of $10$ steps means something very different
	%if $\deltat = 0.1$ and if $\deltat = 0.0001$.
        A rule of thumb
	is that the time horizon of an agent with discount factor $\gamma$ is
	of order $\frac{1}{1 - \gamma}$ steps. If $\gamma_\deltat$ was left constant
	as $\deltat$ goes to $0$, the corresponding time horizon in physical time
	would write as $\frac{\deltat}{1 - \gamma}$ which goes to $0$ when
	$\deltat$ goes to $0$.
        On the contrary, the definition of $\gamma_\deltat$ in
	Eq.~\eqref{eq:def-gamma} yields an approximately constant
	time horizon in physical time when $\deltat$ goes to $0$. Indeed,
	\begin{equation}
          \label{eq:time-horizon}
          \frac{\deltat}{1-\gamma^\deltat}
          = - \frac{1}{\log \gamma} + \bigO(\deltat)
          \approx \frac{1}{1-\gamma},
        \end{equation}
        approximately invariant to $\deltat$ when $\deltat$ goes to 0.

@@

The definition of the return Eq.~\eqref{eq:continuous-return}, yields the
following definition of the continuous value function
\begin{align}
  V^\pi(s) =& \E_{\tau\sim\pi}\left[R(\tau) \mid s_0 = s\right]\\
  =& \E_{\tau\sim\pi}\left[\int\limits_{0}^\infty \gamma^{t} r(s_t, a_t) dt \mid s_0 = s\right].
\end{align}
The corresponding discretization for time discretization $\deltat$ is
\begin{align}
  V^\pi_\deltat(s) =& \E_{\tau\sim\pi}\left[R_\deltat(\tau) \mid s_0 = s\right]\\
  =& \E_{\tau\sim\pi}\left[\sum\limits_{k=0}^\infty \gamma^{k\deltat}r(s^k_{\deltat}, a^k_{\deltat})\deltat \mid s_0 = s\right]
\end{align}
which fits the usual definition of the discrete V-function for $\gamma_\deltat$ and
$r_\deltat$ defined in~(\ref{eq:def-gamma}, \ref{eq:def-reward}).
The discretized value function follows the Bellman equation\footnote{
	If the continuous MDP dynamic follows~\eqref{eq:diffusion}, the limit of the Bellman equation for $V^\pi_\deltat$ when $\deltat \rightarrow 0$ is the Hamilton-Jacobi-Bellman equation on $V^\pi$ \cite{cont_rl}
\begin{equation}
  \label{eq:hamilton-jacobi-bellman}
  r + \partial_s V^\pi \cdot F = - \log(\gamma) V^\pi
\end{equation}}

\begin{equation}
  \label{eq:bellman}
  V^\pi_\deltat(s) = \reward(s, \pi(s))\deltat + \gamma^{\deltat} \E_{s_{\deltat}^{k+1} | s_{\deltat}^{k} = s} V^\pi_\deltat(s_{\deltat}^{k+1})
\end{equation}

We have the following convergence theorem of the discretized value function to the continuous value function:
  \begin{theorem}
   For all $s \in {\cal S}$:
    \begin{equation}
      \label{eq:conv-value}
      V^\pi_\deltat(s) = V^\pi(s) + \bigO(\deltat)
    \end{equation}
    \label{th:conv-value}
  \end{theorem}

@@

Contrary to the Value function, the Action-Value function $Q$ is ill-defined
for a continuous MDP. More precisely, the Q-function  collapses to the
V-function when $\deltat \rightarrow 0$. In near continuous time, the effect of
individual actions on the Q-function is of order $\bigO(\deltat)$. This makes
deriving a ranking of action from an approximate Q-function difficult.
Formally:
\begin{theorem}
	Let $\pi$ be a deterministic policy such that for all $\deltat$, $V^\pi_\deltat$ is continuously
	differentiable, then, for all $s, a$,
	\begin{equation}
		\label{eq:q-discr}
		Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO\left(\deltat\right)
	\end{equation}

  \label{th:q-cont}
\end{theorem}
% Consider a discretization of Eq.~\eqref{eq:cont_bell} with time interval
% $\deltat \ll~1$, Ito's lemma yield
% \begin{align}
% 	Q^\pi(s, a) &= \reward(s, a) \deltat + \gamma^{\deltat} \E_{\delta s \sim \delta s(s, a)}V^\pi(s + \delta s)\nonumber\\
% 		    &= \reward(s, a) \deltat + \gamma^{\deltat} V^\pi(s) +
% 		     \gamma^{\deltat} \partial_s V^\pi(s)f(s, a) \deltat\nonumber\\
% 		    &+ \frac{1}{2}\text{tr}\left(\Sigma^T\partial^2_{s^2}V^\pi(s)\Sigma\right)\deltat + \bigO(\deltat^2)\nonumber\\
% 		    &\xrightarrow[\deltat \to 0]{} V^\pi(s).
% 		    \label{eq:Q_cont}
                      %   \end{align}

A consequence of Theorem \ref{th:q-cont} is that in exact continuous time,
$Q^\pi$ does not bear any information on the ranking of actions, and
thus cannot be used to select actions that yields higher future returns. There is no time continuous Q-learning.

\begin{proof}
	Bellman equation for the discretized Q-function yields
  \begin{align}
    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V^\pi_\deltat(s')\right] 
  \end{align}
  From the environment dynamic equation, 
  \begin{equation}
	  s' = s + f(s, a) \deltat + \smallo(\deltat).
  \end{equation}As $V^\pi$ is assumed to be continuously differentiable,
  \begin{align}
	  V^\pi_\deltat(s') &= V^\pi_\deltat(s) + \partial_s V^\pi_\deltat(s) F(s, a)\deltat + \smallo(\deltat)\\
			    &= V^\pi_\deltat(s) + \bigO(\deltat)
  \end{align}
  Rewriting $Q^\pi_\deltat$ yields
  \begin{align}
	  Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat (V^\pi_\deltat(s) + \bigO(\deltat))\\
			      &= (1 + \bigO(\deltat)) (V^\pi_\deltat(s) + \bigO(\deltat)) + \bigO(\deltat)\\
			      &= V^\pi_\deltat(s) + \bigO(\deltat).
  \end{align}
  Now, using Thm.~\ref{th:conv-value},
  \begin{align}
	  Q^\pi_\deltat(s, a) = V^\pi(s) + \bigO(\deltat)
  \end{align}
  which ends the proof.
%  But we know that $s'$ is close to $s$. If we define $\delta s = s' - s$, then:
%  \begin{equation}
%    \E_{s'|s, a}\left[\delta s\right] = \bigO(\deltat) \quad \text{Var}(\delta s) = \bigO(\deltat)
%  \end{equation}
%  \TODO{Explain. And this is not exactly the Variance.} Therefore:
%  \begin{align}
%    |\E_{s'|s, a}\left[V^\pi_\deltat(s')\right] - V^\pi_\deltat(s)| &= \E_{\delta s|s, a}\left[\partial_s V^\pi_\deltat\cdot \delta s +\bigO(\|\delta s\|^2)\right]\nonumber\\
%                                            &\leq \|\partial_s V_\deltat^\pi\|\cdot \|\E_{\delta s|s, a}\left[\delta s\right]\| +\bigO(\delta t)\nonumber\\
%    &= \bigO(\deltat)\nonumber
%  \end{align}
%  Finally:
%  \begin{align}
%    Q^\pi_\deltat(s, a) &= r(s, a)\deltat + \gamma^\deltat \E_{s'|s, a}\left[V_\deltat^\pi(s')\right] \nonumber\\
%                &= r(s, a)\deltat + \left(1 + \bigO(\deltat)\right)(V^\pi(s) + \bigO(\deltat))\nonumber\\
%    &= V^\pi(s) + \bigO(\deltat)\nonumber
%  \end{align}
  \end{proof}
